---
title: "GeoDengue: Mapping Dengue Risk Across Niterói Neighborhoods"
author: "Camila Totti"
date: "2026-06-18"
output: html_document
---

## Executive Summary

Dengue remains one of the main public health challenges in Brazil. Understanding its spatial distribution is essential for prioritizing surveillance actions, allocating resources, and supporting evidence-based decision making.

This project analyzes reported dengue cases in Niterói (Rio de Janeiro, Brazil) during 2024 and estimates the incidence rate per 100,000 inhabitants for each neighborhood.

## Key Deliverables

- Data cleaning and standardization workflow
- Integration of epidemiological and census datasets
- Calculation of dengue incidence rates
- Neighborhood-level spatial analysis
- Choropleth map highlighting priority areas

------------------------------------------------------------------------

## Business Problem

Public health managers need to identify areas with higher dengue burden in order to:

- Prioritize vector-control actions
- Optimize field team deployment
- Improve resource allocation
- Support epidemiological surveillance

Raw case counts alone may be misleading because neighborhoods have different population sizes.

To address this issue, incidence rates were calculated using population-adjusted indicators.

------------------------------------------------------------------------

## Data Sources

### Dengue Notification Database

Contains reported dengue cases including:

- Notification year
- Neighborhood information
- Administrative identifiers

### Population Data

Neighborhood-level population estimates used to calculate incidence rates.

### Administrative Boundaries

Neighborhood boundaries provided in GeoJSON format.

```{r setup, include=FALSE}
library(tidyr)
library(dplyr)
library(readxl)
library(stringi)
library(stringr)
library(sf)
library(leaflet)
library(ggplot2)

dengue <- read_excel("Banco_dengue_teste.xlsx")
census <- read.csv2("censo2217762015942.csv") %>%
  rename(
    neighborhood = Bairro,
    resident_population = Populacao.residente
  )
```

------------------------------------------------------------------------

## Methodology

The workflow consisted of four main stages:

### 1. Data Cleaning

The dengue database presented:

- Records from municipalities other than Niterói
- Missing neighborhood information
- Inconsistencies in neighborhood names

Examples of corrections applied:

| Original Value | Standardized Value |
|----------------|--------------------|
| FATIMA         | BAIRRO DE FATIMA   |
| PONTA DA AREIA | PONTA D'AREIA      |
| VICOSO JARDIM  | VICOSO JARDIM      |

Neighborhood names were standardized to ensure compatibility across epidemiological, demographic and spatial datasets.

``` r
dengue_24 <- dengue %>%
  filter(NU_ANO == "2024") %>%
  rename(neighborhood = NM_BAIRRO) %>%           
  filter(!is.na(neighborhood)) %>%
  mutate(
    neighborhood = str_squish(neighborhood),
    neighborhood = str_to_upper(neighborhood),
    neighborhood = recode(
      neighborhood,
      "FATIMA" = "BAIRRO DE FATIMA",
      "PONTA DA AREIA" = "PONTA D'AREIA",
      "VICOSO  JARDIM" = "VICOSO JARDIM"
    )
  )
```

```{r processing, echo=FALSE, message=FALSE, warning=FALSE}
dengue_24 <- 
  dengue %>%
  filter(NU_ANO == "2024") %>%
  rename(neighborhood = NM_BAIRRO) %>%
  filter(!is.na(neighborhood)) %>%
  mutate(
    neighborhood = str_squish(neighborhood),
    neighborhood = str_to_upper(neighborhood),
    neighborhood = recode(
      neighborhood,
      "FATIMA" = "BAIRRO DE FATIMA",
      "PONTA DA AREIA" = "PONTA D'AREIA",
      "VICOSO  JARDIM" = "VICOSO JARDIM"
    )
  ) %>%
  group_by(neighborhood) %>%
  summarise(
     reported_cases = n(),
    .groups = "drop"
  )

census <-
  census %>%
  filter(neighborhood != "Total") %>%
  mutate(
    neighborhood = str_squish(neighborhood),
    neighborhood = str_to_upper(neighborhood),
    neighborhood = str_replace_all(neighborhood, "\\\\'", "'")
  )

dengue_by_neighborhood <-
  census %>%
  left_join(
    dengue_24,
    by = "neighborhood"
  ) %>%
  mutate(
    reported_cases = coalesce(reported_cases, 0),
    incidence_per_100k =
      round(
        (reported_cases / resident_population) * 100000,
        2
      )
  )
```

### 2. Case Aggregation

Cases were grouped by neighborhood to estimate the total number of notifications.

### 3. Population Integration

The dengue dataset was joined with census information.

### 4. Incidence Rate Calculation

The incidence rate was calculated as:

$$Incidence = \frac{Reported\ Cases}{Population} \times 100,000$$

------------------------------------------------------------------------

### Spatial Analysis

Neighborhood boundaries were integrated with epidemiological indicators through a spatial join.

```{r spatial_processing, echo=FALSE}
neighborhoods <- st_read(
  "Limite_de_Bairros.geojson",
  quiet = TRUE
)

neighborhoods <- neighborhoods %>%
  mutate(
    neighborhood_std =
      stri_trans_general(tx_nome, "Latin-ASCII") %>%
      toupper() %>%
      str_trim()
  )

dengue_map <- neighborhoods %>%
  left_join(
    dengue_by_neighborhood,
    by = c(
      "neighborhood_std" = "neighborhood"
    )
  )
```

------------------------------------------------------------------------

## Results

### Dengue Incidence by Neighborhood

#### Top 5 Neighborhoods by Incidence Rate

```{r top5_plot, echo=FALSE, message=FALSE, warning=FALSE, fig.width=8, fig.height=4}
library(plotly)

top_n <- 10

df_plot <- dengue_by_neighborhood %>%
  arrange(desc(incidence_per_100k)) %>%
  slice_head(n = top_n)

p <- ggplot(df_plot, aes(
  x = reorder(neighborhood, incidence_per_100k),
  y = incidence_per_100k,
  text = paste0(
    "Neighborhood: ", neighborhood, "<br>",
    "Incidence: ", round(incidence_per_100k, 1), "<br>",
    "Cases: ", reported_cases
  )
)) +
  geom_col(fill = "#991766") +
  coord_flip() +
  labs(x = NULL, y = NULL) +
  theme_minimal()

ggplotly(p, tooltip = "text")
```

Higher incidence rates were concentrated in a subset of neighborhoods, while several areas presented comparatively lower levels of dengue burden.

```{r fig.width=9, fig.height=7, echo=FALSE}
library(leaflet)
library(RColorBrewer)
library(htmltools)

pal <- colorNumeric(
  palette = colorRampPalette(c("#fdf2c5", "#ff6e27", "#991766", "#331327"))(256),
  domain = dengue_map$incidence_per_100k,
  na.color = "transparent"
)

dengue_map <- dengue_map %>%
  mutate(
    incidence_label = ifelse(
      is.na(incidence_per_100k),
      paste0(
        "<b>", tx_nome, "</b><br>",
        "Incidence: NA<br>",
        "Cases: ", ifelse(is.na(reported_cases), 0, reported_cases), "<br>",
        "Population: ", resident_population
      ),
      paste0(
        "<b>", tx_nome, "</b><br>",
        "Incidence: ", round(incidence_per_100k, 1), " per 100k<br>",
        "Cases: ", reported_cases, "<br>",
        "Population: ", resident_population
      )
    )
  )


leaflet(dengue_map) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  
  addPolygons(
    fillColor = ~pal(incidence_per_100k),
    fillOpacity = 0.85,
    color = "white",
    weight = 1,
    
    label = ~lapply(incidence_label, HTML),
    
    highlightOptions = highlightOptions(
      weight = 2,
      color = "#333",
      bringToFront = TRUE
    )
  ) %>%
  
  addLegend(
  pal = pal,
  values = dengue_map$incidence_per_100k,
  title = "Dengue incidence<br>(per 100k)",
  position = "bottomright",
  na.label = " "
) 

```

```{r executive_metrics, echo=FALSE, results='asis'}
metrics <- dengue_by_neighborhood %>%
  summarise(
    neighborhoods = n(),
    total_cases = sum(reported_cases, na.rm = TRUE),
    avg_incidence = round(mean(incidence_per_100k, na.rm = TRUE), 1),
    median_incidence = round(median(incidence_per_100k, na.rm = TRUE), 1),
    max_incidence = round(max(incidence_per_100k, na.rm = TRUE), 1)
  )

cat(sprintf('
<div style="display:flex; flex-wrap:wrap; gap:15px;">

<div style="padding:15px; border-radius:10px; background:#f5f5f5; width:180px;">
<h4>Neighborhoods</h4>
<h2>%s</h2>
<p>Analyzed</p>
</div>

<div style="padding:15px; border-radius:10px; background:#f5f5f5; width:180px;">
<h4>Cases</h4>
<h2>%s</h2>
<p>Reported</p>
</div>

<div style="padding:15px; border-radius:10px; background:#f5f5f5; width:180px;">
<h4>Mean Incidence</h4>
<h2>%s</h2>
<p>per 100k</p>
</div>

<div style="padding:15px; border-radius:10px; background:#f5f5f5; width:180px;">
<h4>Median Incidence</h4>
<h2>%s</h2>
<p>per 100k</p>
</div>

<div style="padding:15px; border-radius:10px; background:#f5f5f5; width:180px;">
<h4>Highest Incidence</h4>
<h2>%s</h2>
<p>per 100k</p>
</div>

</div>',
metrics$neighborhoods,
format(metrics$total_cases, big.mark = ","),
metrics$avg_incidence,
metrics$median_incidence,
metrics$max_incidence
))

```

------------------------------------------------------------------------

## Key Findings

- A total of 51 neighborhoods were analyzed across Niterói.
- Dengue incidence ranged from 0 to 2,991 cases per 100,000 inhabitants, revealing substantial spatial heterogeneity across the municipality.
- The median incidence rate was 477 cases per 100,000 inhabitants, indicating that half of the neighborhoods experienced incidence rates above this level.
- The average incidence rate reached 587 cases per 100,000 inhabitants, suggesting that a subset of neighborhoods with very high incidence rates influenced the municipal average.
- Several neighborhoods exhibited incidence rates significantly above the municipal median, highlighting potential priority areas for surveillance and vector-control interventions.

### Limitations

- One neighborhood ("Vital Brazil") could not be evaluated due to missing population information in the census dataset.

------------------------------------------------------------------------

## Skills Demonstrated

- Data Cleaning and Standardization
- Data Integration
- Spatial Data Processing
- Epidemiological Indicators
- Interactive Web Mapping
- Data Visualization
- Reproducible Analytics using R Markdown
